---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true

author: "Lisa Rodgers"
date: "May 9, 2016"
output: html_document
---

#Analysis of personal activity monitoring device data

-----


##Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

In this report we make use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day. Data was provided for the course students for this particular assignment, avaliable to download via this link (avaliable at 2014-05-25).

In the report, we try to answer the following questions:

* What is mean total number of steps taken per day?
* What is the average daily activity pattern?
* Are there differences in activity patterns between weekdays and weekends?


-----

## Loading and preprocessing the data

-----

1. Load the data (i.e. read.csv())

Load the necessary libraries

```{r loadLibrary}
library(ggplot2)
```


```{r loadFiles}
if(!file.exists('activity.csv')){
    unzip('activity.zip')
}
activityData <- read.csv('activity.csv', header = TRUE, sep = ",", na.strings = "NA")
```




Looking at a summary for the dataset using “summary” and “str” methods:

```{r summaryData}
summary(activityData)
```

```{r DataStructure}
str(activityData)
```

```{r TopofData}
head(activityData)
```

```{r BottomofData}
tail(activityData)
```

##### 2. Process/transform the data (if necessary) into a format suitable for your analysis
Convert the “date” variable to a Date class and the “interval” variable to a factor

```{r ConvertDate}
activityData$date <- as.Date(activityData$date, format = "%Y-%m-%d")
activityData$interval <- factor(activityData$interval)
```

-----

### What is mean total number of steps taken per day?

-----

1 - Subsetting the dataset to ignore missing values


```{r RemoveNulls}
NA_index <- is.na(as.character(activityData$steps))
activityData_noNull<- activityData[!NA_index,]
head(activityData_noNull)
```



2 - Aggregating the number of steps taken each day

Creating a data frame with the steps taken for each day

```{r AggregateData}
steps_each_day <- aggregate(steps ~ date, data = activityData_noNull, sum)
```

Adding column names to the created data frame

```{r ColumnNames}
colnames(steps_each_day) <- c("date", "steps")
```

3 - Making a histogram of the total number of steps taken each day:
```{r PlotTotalSteps}
hist(as.numeric(steps_each_day$steps), breaks = 20, col = "blue", xlab = "Number of Steps", main= "Histogram of the Daily Steps")
```

median total number of steps taken per day:

**Mean**
```{r GetMeanSteps}
mean(steps_each_day$steps)
```


**Median**
```{r GetMedianSteps}
median(steps_each_day$steps)
```

-----

##What is the average daily activity pattern?

-----

1 - Calculating the average number of steps taken, averaged across all days:

Calculating the average
```{r StepPerInterval}
steps_per_interval <- aggregate(activityData_noNull$steps, by=list(interval=activityData_noNull$interval), FUN=mean)
```

Adding columns names
```{r AddColumnNames}
colnames(steps_per_interval) <- c("interval", "average_steps")
```

Ploting the average daily activity pattern 
```{r AverageDailyPatterns}
plot(as.integer(levels(steps_per_interval$interval)), steps_per_interval$average_steps, type="l",
     xlab = "Interval", ylab = "Average Number of Steps", main = "Average Daily Activity Pattern",  col ="blue")

```

2 - The 5-minute interval that contains the maximum number of steps:

```{r max_steps}
max_steps <- max(steps_per_interval$average_steps)

```
The maximum number of average steps was `r max_steps`.

The 5-minute interval that contains the maximum number of `r max_steps` steps.
```{r IntervalMaxSteps}
intervale_max_steps<-steps_per_interval[which.max(steps_per_interval$average_steps),]$interval

```
The 5-minute interval that contains the maximum number of steps `r intervale_max_steps`


288 Levels: 0 5 10 15 20 25 30 35 40 45 50 55 100 105 110 115 120 ... 2355

So, the 5-minute interval that contains the maximum number of steps (`r round(max_steps)`) is the interval `r intervale_max_steps`.

Inputing missing values
1 - The total number of missing values in the dataset (for each variable):

For the “steps” variable:
```{r NullSteps}
NA_Steps <- sum(is.na(as.character(activityData$steps)))
```

Total number of steps that were NA: **`r NA_Steps`** 

For the “date” variable:
```{r NullDates}
NA_Date <-sum(is.na(as.character(activityData$date)))
```

Total number of Dates that were NA: **`r NA_Date`** 


For the “interval” variable:
```{r NullIntervals}
NA_Intervals <-sum(is.na(as.character(activityData$interval)))
```
Total number of Intervals that were NA: **`r NA_Intervals`**

Therefore, the total number of missing values in the dataset is **`r NA_Steps + NA_Date + NA_Intervals`**.

2- The strategy for filling in all of the missing values in the dataset. Missing values are replaced by the mean of that 5-minute interval.

finding the indices of missing values (NAs)
```{r NullIndex}
NA_index <- which(is.na(as.character(activityData$steps)))
complete_data <- activityData
```

Inputing missing values using the mean for that 5-minute interval
```{r CompleteData}
complete_data[NA_index, ]$steps<-unlist(lapply(NA_index, FUN=function(NA_index){
                steps_per_interval[activityData[NA_index,]$interval==steps_per_interval$interval,]$average_steps
                }))
```

3 - Creating a new dataset that is equal to the original dataset but with the missing data filled in.

Checking the complete data with the summary and str methods
```{r SummaryCompleteData}
summary(complete_data)

str(complete_data)
```

4 - Making a histogram of the total number of steps taken each day for the complete dataset:

Creating a data frame with the steps taken for each day
```{r Steps_Completed_Each_Day}
steps_each_day_complete <- aggregate(steps ~ date, data = complete_data, sum)
```


Adding column names to the created data frame
```{r}
colnames(steps_each_day_complete) <- c("date", "steps")
```

Making the histogram
```{r PlotTotalStepsEachDay}
hist(as.numeric(steps_each_day_complete$steps), breaks = 20, col = "red", xlab = "Number of Steps", main= "Total Number of Steps Taken Each Day")
```

